---
title: "isolated_HGS_ROSMAP_models"
author: "Rachel Bercovitch"
date: "4/23/2023"
output: html_document
---

## LOAD PACKAGES
```{r include=FALSE}
library(Hmisc)
library(dplyr)
library(corrplot)
library(tidyr)
library(tidyverse)
library(reshape2)
library(ggplot2)
library(cowplot)
library(plyr)
library(magrittr)
library(lme4)
library(lmerTest)
library(rms)
library(mediation)
library(tibble)
library(readxl)
library(readr)  
library(gtools)
```

## LOAD DATA
```{r}
rosmap_path<-'/external/rprshnas01/netdata_kcni/dflab/team/rb/ROSMAP/' # file path on SCC

# cross-sectional

ROSMAP_cross <- read.csv(file.path(rosmap_path,'ROSMAP_basic_04-21-2023_noDup.csv'))
ROSMAP_cross <- ROSMAP_cross[,c("projid", "age_bl", "msex", "educ", "age_first_ad_dx", "age_death")]

# longitudinal

long_xl <- read_excel(file.path(rosmap_path,'dataset_978_long_04-21-2023.xlsx'))
write_csv(long_xl, file="/external/rprshnas01/netdata_kcni/dflab/team/rb/ROSMAP/ROSMAP_long_04-21-2023.csv")
ROSMAP_long <- read.csv(file.path(rosmap_path,'ROSMAP_long_04-21-2023.csv'))
ROSMAP_long <- ROSMAP_long[,c("projid", "vasc_risks_sum", "phys5itemsum", "fu_year", "mot_grip", "d_frailty", "dcfdx", "age_at_visit", "bmi", "cogn_global")]

# post-mortem

post <- readRDS("/external/rprshnas01/external_data/rosmap/metadata/ROSmaster.rds")
ROSMAP_post <- post[,c("projid", "amyloid", "tangles", "nft", "pmi", "cogdx", "plaq_n", "plaq_d", "tdp_stage4", "cvda", "arteriol_scler", "cvda_4gp2", "ci_num2_gct", "ci_num2_mct", "caa_4gp", "dlbdx")]

```

## FORMAT LONGITUDINAL DATA
```{r}
## Participant subset to calculate cognitive slopes for in longitudinal analysis

ID_subset <- names(table(ROSMAP_long$projid)>1) # subset the data upstream to people who have >2 visits 
pl <- subset(ROSMAP_long, projid %in% ID_subset)

## Get average score for physical activity, vascular risk, and BMI

mean_obs <- lapply(unique(pl$projid),function(i) {
  
  data_sub <- subset(pl, projid==i) # subset pl dataset to each person (i) one by one
  data_sub$phys_mean <- mean(data_sub[,"phys5itemsum"], na.rm=T) # average over all visits
  data_sub$vasc_mean <- mean(data_sub[,"vasc_risks_sum"], na.rm=T) # average over all visits
  data_sub$bmi_mean <- mean(data_sub[,"bmi"], na.rm=T) # average over all visits
  data_sub[1,c("projid","phys_mean","vasc_mean", "bmi_mean")] 
  
})

mean_df <- as.data.frame(do.call(rbind,mean_obs)) # data frame with three avg. columns to merge into cross-sectional data and PRS data
```

## CALCULATE COGNITION SLOPE FOR EACH PARTICIPANT
```{r}
## Linear mixed effects model to obtain change in cognition over time for each participant

longitudinal_rosmap <- readRDS("/external/rprshnas01/netdata_kcni/dflab/team/rb/rb-rosmap-plink-profile-merge/data/data-merge/data_merge_nosqldf.RDS")
long_model3 <- lmer(cogn_global ~ age_bl + fu_year + (1 + fu_year|projid), data = longitudinal_rosmap, REML = FALSE,control = lmerControl(optimizer ="Nelder_Mead"))
coefs <- coef(long_model3)[["projid"]]
fu_year_coefs <- coefs[3] 
cog_change_over_time <- tibble::rownames_to_column(fu_year_coefs, "projid") # re-introduce projid as column name
colnames(cog_change_over_time)[2] <- "fu_year_coefs"
```

## OBTAIN CROSS-SECTIONAL COGNITION VARIABLE
```{r}
## Get the most recent global cognition score

cognition_ROSMAP_long <- subset(ROSMAP_long, is.na(cogn_global)==FALSE)
cog_last_visit <- cognition_ROSMAP_long %>%
  group_by(projid) %>%
  slice_max(order_by = age_at_visit, n = 1) # order participants by age at visit and take the max corresponding value for cognition
cog_last_visit$cogn_global_last_visit <- cog_last_visit$cogn_global
cog_last_visit$age_last_cogn_test <- cog_last_visit$age_at_visit
cog_last_visit <- cog_last_visit[,c("cogn_global_last_visit", "age_last_cogn_test", "projid")]

# Get the most recent grip strength measurement

handstrength_ROSMAP_long <- subset(ROSMAP_long, is.na(mot_grip)==F)
handstreng_last_visit <- handstrength_ROSMAP_long %>%
  group_by(projid) %>%
  slice_max(order_by = age_at_visit, n = 1) # order participants by age at visit and take the max corresponding value for grip strength
handstreng_last_visit$handstrength_last_visit <- handstreng_last_visit$mot_grip
handstreng_last_visit$age_last_handstrength_test <- handstreng_last_visit$age_at_visit
handstreng_last_visit <- handstreng_last_visit[,c("handstrength_last_visit", "age_last_handstrength_test", "projid")]

last_visit_df <- merge(cog_last_visit, handstreng_last_visit, by = "projid") # merge these two new variables (last global cognition and last grip strength) into one df
```

## PRS DATA
```{r}
## Import PRS and merge them to wide-form df
# PRS from Jones HGS GWAS

HGS_PRS <- read.table("/external/rprshnas01/netdata_kcni/dflab/team/rw/projects/rb-rosmap-pca/data/03_features/rosmap/polygenic-risk-score/hand-grip-strength/regressed_polygenic_risk_score.tsv", header = T)
colnames(HGS_PRS)[2] <- "HGS_PRS"
HGS_PRS$HGS_PRS <- (HGS_PRS$HGS_PRS)*(-1) # multiply by -1 to aid in interpretation (grip strength vs. grip weakness)
colnames(HGS_PRS)[1] <- "projid"

## Construct PRS-AD

AD_PRS <- read.table("/external/rprshnas01/netdata_kcni/dflab/team/rw/projects/rb-rosmap-pca/data/03_features/rosmap/polygenic-risk-score/alzheimer-disease/regressed_polygenic_risk_score.tsv", header = T)
colnames(AD_PRS)[2] <- "AD_PRS"
colnames(AD_PRS)[1] <- "projid"

PRS_df <- merge(HGS_PRS, AD_PRS, by = "projid") # merge PRS to one df
```

## MERGING ALL DATASETS
```{r}
ROSMAP_post$projid <- as.integer(ROSMAP_post$projid)
cog_change_over_time$projid <- as.integer(cog_change_over_time$projid)
df_list <- list(ROSMAP_cross, ROSMAP_post, mean_df, cog_change_over_time, PRS_df, last_visit_df)
ROSMAP_df <- df_list %>% reduce(full_join, by = "projid")
```

## DEMOGRAPHIC INFORMATION
```{r}
# Stratify ROSMAP_df by sex

male.df <- ROSMAP_df[which(ROSMAP_df$msex == "1"),]
female.df <- ROSMAP_df[which(ROSMAP_df$msex == "0"),]
summary_vars <- c("age_bl", "bmi_mean", "phys_mean", "vasc_mean", "cogdx")

demo_stats <- for (i in summary_vars) {
  
  cat(paste0("Variable: ", i, "\n"))
  cat("Male summary statistics:\n")
  print(summary(male.df[[i]]))
  
  cat("\nFemale summary statistics:\n")
  print(summary(female.df[[i]]))
  
  cat("\n")
  
}

```

## SCALE NUMERIC VARS
```{r}
# List of variable names to scale and convert

vars_to_scale <- c("age_bl", "phys_mean", "bmi_mean", "HGS_PRS", "AD_PRS", "cogn_global_last_visit", "age_last_cogn_test","handstrength_last_visit","age_last_handstrength_test")

# For loop to scale and convert to numeric
for (i in vars_to_scale) {
  ROSMAP_df[[i]] <- as.numeric(scale(ROSMAP_df[[i]]))
}

# square root transform brain pathology variables

vars_to_transform <- c("amyloid", "tangles", "nft", "plaq_d", "plaq_n")

# For loop to scale and square root transform

for (i in vars_to_transform) {
  ROSMAP_df[[i]] <- scale(sqrt(ROSMAP_df[[i]]))
}

```

## VALIDATION MODELS
```{r}
# validate our PRS-HGS with measured hand strength

validate_HGS <- lm(handstrength_last_visit~HGS_PRS, data = ROSMAP_df)
validate_HGS.2 <- lm(handstrength_last_visit~HGS_PRS + age_last_handstrength_test, data = ROSMAP_df)
validate_HGS.3 <- lm(handstrength_last_visit~HGS_PRS + msex, data = ROSMAP_df)
validate_HGS_covs <- ols(handstrength_last_visit~HGS_PRS + age_last_handstrength_test + msex + educ + bmi_mean, data = ROSMAP_df)
```

## MODEL PRS-HGS WITH RESPECT TO MOST RECENT GLOBAL COGNITION SCORE
```{r}
CS_model_1a <- lm(cogn_global_last_visit~HGS_PRS, data = ROSMAP_df)
CS_model_1b <- ols(cogn_global_last_visit~HGS_PRS + age_last_cogn_test + msex + phys_mean + vasc_mean + educ + bmi_mean, data = ROSMAP_df)
CS_model_1c <- lm(cogn_global_last_visit~AD_PRS + age_last_cogn_test + msex + phys_mean + vasc_mean + educ + bmi_mean, data = ROSMAP_df)
CS_model_1d <- lm(cogn_global_last_visit~AD_PRS + HGS_PRS + age_last_cogn_test + msex + phys_mean + vasc_mean + educ + bmi_mean, data = ROSMAP_df)
```

## LONGITUDINAL LINEAR MODELS (cognition~PRS-HGS)
```{r}
# hypothesis 1 

model_1a.i <- lm(fu_year_coefs~HGS_PRS, data = ROSMAP_df)
model_1a.ii <- ols(fu_year_coefs~HGS_PRS + age_bl + msex + phys_mean + vasc_mean + educ + bmi_mean, data = ROSMAP_df)

# hypothesis 2

model_2.1 <- lm(fu_year_coefs~AD_PRS, data = ROSMAP_df)
model_2.2 <- lm(fu_year_coefs~AD_PRS + age_bl + msex + phys_mean + vasc_mean + bmi_mean + educ, data = ROSMAP_df)
model_2.3 <- lm(fu_year_coefs~AD_PRS + HGS_PRS + age_bl + msex + phys_mean + vasc_mean + bmi_mean + educ, data = ROSMAP_df)
model_2.4 <- lm(fu_year_coefs~AD_PRS * HGS_PRS + age_bl + msex + phys_mean + vasc_mean + bmi_mean + educ, data = ROSMAP_df)

# p.adjust(c(mod2sum_1$coefficients[2,4], mod2sum_2$coefficients[2,4], mod2sum_3$coefficients[2,4]), method = "fdr")
```

## POST-MORTEM MODELS
```{r}

# structure of models:
# model_pathology.1 = lm(pathology~AD_PRS + handstrength_last_visit + msex + age_death + pmi, data = ROSMAP_df)
# model_pathology.2 = lm(amyloid~HGS_PRS + handstrength_last_visit + msex + age_death + pmi, data = ROSMAP_df)
# model_pathology.3 = lm(amyloid~handstrength_last_visit + msex + age_death + pmi, data = ROSMAP_df)

model_list <- list() # list to store the models

# vector of dependent variables
dependent_vars <- c("amyloid", "tangles", "nft", "tdp_stage4", "arteriol_scler", "cvda_4gp2", "ci_num2_gct", "ci_num2_mct", "plaq_d", "plaq_n")

# vector of independent variables
independent_vars <- c("AD_PRS", "HGS_PRS", "handstrength_last_visit", "msex", "age_death", "pmi")

# Loop over the dependent variables
for(var in dependent_vars){
  # Loop over the independent variables
  for(i in 1:length(independent_vars)){
    # Create the model formula
    formula <- paste(var, "~", paste(independent_vars[1:i], collapse = "+"))
    # Create the model
    model <- lm(formula, data = ROSMAP_df)
    # Add the model to the list
    model_list[[paste0("model_",var,".",i)]] <- model
  }
}

```

## MEDIATION ANALYSIS
```{r}
# mediator = physical activity

phys_mediation_df <- ROSMAP_df[,c("phys_mean", "HGS_PRS", "age_last_cogn_test", "educ", "msex", "vasc_mean", "bmi_mean", "cogn_global_last_visit")]
phys_mediation_df <- na.omit(phys_mediation_df)
PA_rosmap_mediate_model <- lm(phys_mean~HGS_PRS + age_last_cogn_test + educ + msex + vasc_mean + bmi_mean, data = phys_mediation_df)
summary(PA_rosmap_mediate_model)
PA_rosmap_full_model <- lm(cogn_global_last_visit~phys_mean + HGS_PRS + age_last_cogn_test + educ + msex + vasc_mean + bmi_mean, data = phys_mediation_df)
set.seed(2000)
PA_rosmap_results <- mediate(PA_rosmap_mediate_model, PA_rosmap_full_model, treat = 'HGS_PRS', mediator = 'phys_mean', boot = T)
summary(PA_rosmap_results)
PA_plot <- plot(PA_rosmap_results)
ggsave(file.path(rosmap_path,'PA_ROSMAP_plot.png'), plot = PA_plot, width = 8, height = 4)

# mediator = hand grip strength

HGS_mediation_df <- ROSMAP_df[,c("HGS_PRS", "age_last_cogn_test", "educ", "msex", "vasc_mean", "bmi_mean", "cogn_global_last_visit", "handstrength_last_visit")]
HGS_mediation_df <- na.omit(HGS_mediation_df)
HGS_rosmap_mediate_model <- lm(handstrength_last_visit~HGS_PRS + age_last_cogn_test + educ + msex + vasc_mean + bmi_mean, data = ROSMAP_df)
summary(HGS_rosmap_mediate_model)
HGS_rosmap_full_model <- lm(cogn_global_last_visit~handstrength_last_visit + HGS_PRS + age_last_cogn_test + educ + msex + vasc_mean + bmi_mean, data = HGS_mediation_df)
set.seed(2000)
HGS_rosmap_results <- mediate(HGS_rosmap_mediate_model, HGS_rosmap_full_model, treat = 'HGS_PRS', mediator = 'handstrength_last_visit', boot = T)
summary(HGS_rosmap_results)
HGS_plot <- plot(HGS_rosmap_results)
ggsave(file.path(rosmap_path,'HGS_ROSMAP_plot.png'), plot = HGS_plot, width = 8, height = 4)
```

## PLOT COGNITIVE DECLINE BY TERTILES
```{r include=FALSE}
rm2 <- subset(longitudinal_rosmap, is.na(cogn_global)==F & is.na(fu_year)==F)
modnew <- lmer(cogn_global ~ age_bl + fu_year + (1 + fu_year|projid), data = rm2, REML = FALSE,control = lmerControl(optimizer ="Nelder_Mead"))
rm2$predicted <- predict(modnew)
rm3 <- rm2
rm3 <- merge(rm3,HGS_PRS,by="projid")
rm2 <- merge(rm2,HGS_PRS,by="projid")
rm2$prs3 <- quantcut(rm2$HGS_PRS,3)
PRS_tertiles <- ggplot(rm2,aes(y=cogn_global,x=fu_year,group=projid))+
  geom_smooth(aes(y=predicted, fill=as.factor(prs3), group=prs3), method="lm")+
  scale_fill_manual(values=c( "red", "#9966CC",  "green"))+
  theme_minimal() +
  labs(x = "Time (years)", y = "Cognition", fill = "PRS-HGS")

ggsave(file.path(rosmap_path,'PRS_tertiles_plot.png'), plot = PRS_tertiles, width = 8, height = 6)
```
